---
title: 'Impact of Weather Events on Public Health and Economy in the United States'
author: "Michał Maliszewski"
date: "16 stycznia 2016"
output: 
  html_document: 
    keep_md: yes
---

## Synopsis

In this report, the goal is to analyze the impact of different weather events on public health and economy based on the storm database collected from the U.S. National Oceanic and Atmospheric Administration’s (NOAA) during the years 1950 - 2012. The data used will be estimates of fatalities, injuries, property, crop and total damage to decide which types of event are most harmful to the population health and economy. My results shows that:

* Heat, Tornado and Flood cause the most deaths.
* Tornado, Lighting and Flood cause the most injuries.
* Flood, Wind and Storm have the highest economical impact.

## Questions

Your data analysis must address the following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

## Data Processing

Our first step will be to load the *.csv file. I assume that the data were before downloaded and unziped. In case of problems with getting the data, it can be downloaded from my GitHub repository (unziped) : 

```{r, echo=TRUE, cache=TRUE}
#Reading the *.csv
rawStormData <- read.csv("repdata-data-StormData.csv")

#Check the number of attributes and rows
dim(rawStormData)
```

To simplify our data set I'm creating additional attribiute which will be Year Number from BGN_DATE attribute.

```{r, echo=TRUE, cache=TRUE}
#Adding year attribuite which is equals to BGN_DATE year part
rawStormData$year <- as.numeric(format(as.Date(rawStormData$BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), "%Y"))

```

Now it's time to plot the histogram of raw data set.

```{r, echo=TRUE}
#Draw a histogram
hist(rawStormData$year, main="Histogram of the total number of events", xlab="Years", col="orange", breaks=seq(from=1950, to=2012, by=2))

```

As we can see, the number of event have been growing since 1990. The highest values start from 1995, however value from 2011 can suggest that we achieved local maximum of events. In case of that, I decided to take to next step only years > 1990. Additionally we don't need to use all the data attribiutes in rawStormData set.
```{r, echo=TRUE, cache=TRUE}
#Takes the data only in [1991 - 2012]
shortStormData <- rawStormData[rawStormData$year > 1990, ]

#Select necessary attributes
neededColumns <- c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP", "year")

#Reduce the number of the data
reducedStormData <- shortStormData[, neededColumns]

#Rename attributes
names(reducedStormData) <- c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP", "YEAR")

#See the result
head(reducedStormData)
```

Now it is time to group the event attribiute.
```{r, echo=TRUE, cache=TRUE}
#Find unique values
uniqueValues <- unique(reducedStormData$EVTYPE)

#Result
head(uniqueValues)
```

So there are 985 unique values.I decided to group them into a 10, most popular (in my opinion) categories. The new attribute will be created.
```{r, echo=TRUE, cache=TRUE}
#Grouping by key words. Consider, that the order of this operations is important
reducedStormData[grepl("precipitation|rain|hail|drizzle|wet|percip|burst|depression|fog|wall cloud", reducedStormData$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Precipitation & Fog"
reducedStormData[grepl("wind|storm|wnd|hurricane|typhoon", reducedStormData$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Wind & Storm"
reducedStormData[grepl("slide|erosion|slump", reducedStormData$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Landslide & Erosion"
reducedStormData[grepl("warmth|warm|heat|dry|hot|drought|thermia|temperature record|record temperature|record high", reducedStormData$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Heat & Drought"
reducedStormData[grepl("cold|cool|ice|icy|frost|freeze|snow|winter|wintry|wintery|blizzard|chill|freezing|avalanche|glaze|sleet", reducedStormData$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Cold & Snow"
reducedStormData[grepl("flood|surf|blow-out|swells|fld|dam break", reducedStormData$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Flood & Surf"
reducedStormData[grepl("seas|high water|tide|tsunami|wave|current|marine|drowning", reducedStormData$EVTYPE, ignore.case = TRUE), "damageSource"] <- "High Sea"
reducedStormData[grepl("tstm|thunderstorm|lightning", reducedStormData$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Lightning"
reducedStormData[grepl("tornado|spout|funnel|whirlwind", reducedStormData$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Tornado"
reducedStormData[grepl("fire|smoke|volcanic", reducedStormData$EVTYPE, ignore.case = TRUE), "damageSource"] <- "Fire & Volcano"
reducedStormData <- reducedStormData[complete.cases(reducedStormData[, "damageSource"]), ]

#Rename attributes
names(reducedStormData) <- c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP", "YEAR", "DMG_SRC")

#Change to a factor for future use
reducedStormData$DMG_SRC <- as.factor(reducedStormData$DMG_SRC)
```

## Impact on population health

Before we start calculations, some sort function will be needed (it will additionally sum the values). I wrote one:
```{r, echo=TRUE, cache=TRUE}
sorter <- function(fieldName, dataset = stormData) {
     index <- which(colnames(dataset) == fieldName)
     field <- aggregate(dataset[, index], by = list(dataset$DMG_SRC), FUN = "sum")
     names(field) <- c("EVTYPE", fieldName)
     field <- arrange(field, field[, 2], decreasing = T)
     field <- head(field, n = 10)
     field <- within(field, EVTYPE <- factor(x = EVTYPE, levels = field$EVTYPE))
     return(field)
}
```
Now we can use it to calculate injuries and deaths depends on event type.
```{r, echo=TRUE, cache=TRUE}
#Plyr library will be needed to use arrange function
library(plyr)

#Counting fatalities
fatalities <- sorter("FATALITIES", dataset = reducedStormData)

#Counting injuries
injuries <- sorter("INJURIES", dataset = reducedStormData)
```

## Impact on economy 

In this step we will count total amount of damages. To do so we have to change the multiplicators to the number format. To achieved it, I had created two functions.
```{r, echo=TRUE, cache=TRUE}

#This function replaces literals to corresponding number for pow function
toTenPower <- function(x){
     if(is.numeric(x)) {
         x <- x
     }
     else if(grepl("h", x, ignore.case=TRUE)) {
         x <- 2
     }
     else if(grepl("k", x, ignore.case=TRUE)) {
         x <- 3
     }
     else if(grepl("m", x, ignore.case=TRUE)) {
         x <- 6
     }
     else if(grepl("b", x, ignore.case=TRUE)) {
         x <- 9
     }
     else{
         x <- 0
     }
     x
}

#Function to calculate amount basing on pow and basic number.
calculateAmount <- function(num, exp){
     pow <- toTenPower(exp)
     if(is.numeric(num)){
         num <- num * (10 ^ pow)
     }
     
     if(!is.numeric(num)){
         num <- 0
     }
     num
 }
```
Now we will use this two functions to create attributes with total property and crop damage.
```{r, echo=TRUE, cache=TRUE}
#Property damage calculation
reducedStormData$PROP_DMG <- mapply(calculateAmount, reducedStormData$PROPDMG, reducedStormData$PROPDMGEXP)

#Crop damage calculation
reducedStormData$CROP_DMG <- mapply(calculateAmount, reducedStormData$CROPDMG, reducedStormData$CROPDMGEXP)

#Total damage calculation
reducedStormData$TOTAL_DMG = reducedStormData$PROP_DMG + reducedStormData$CROP_DMG
```

Additionally we created a total damage amount which is a sum of crop and property damge. The last step is to aggregate and ordem them.
```{r, echo=TRUE}
#Aggregate the data
cropDmg <- aggregate(reducedStormData$CROP_DMG~DMG_SRC, data=reducedStormData, sum)

#Change names
names(cropDmg) <- c("EVENT", "CROP_DMG")

#Set in right order
cropDmg <- cropDmg[order(-cropDmg$CROP_DMG),]

```
The same 3 steps will be done for property and total damage:
```{r, echo=TRUE}
#Aggregate the data
propDmg <- aggregate(reducedStormData$PROP_DMG~DMG_SRC, data=reducedStormData, sum)

#Change names
names(propDmg) <- c("EVENT", "PROP_DMG")

#Set in right order
propDmg <- propDmg[order(-propDmg$PROP_DMG),]

```

```{r, echo=TRUE}
#Aggregate the data
totalDmg <- aggregate(reducedStormData$TOTAL_DMG~DMG_SRC, data=reducedStormData, sum)

#Change names
names(totalDmg) <- c("EVENT", "TOTAL_DMG")

#Set in right order
totalDmg <- totalDmg[order(-totalDmg$TOTAL_DMG),]

```

## Results

To answer the first question, we will use an injuries and fatalities data, which was prepared before. 
```{r, echo=TRUE, cache=TRUE}
injuries
```
```{r, echo=TRUE, cache=TRUE}
fatalities
```
The results can be shown on the plots below.
```{r, echo=TRUE, fig.keep="none"}
#Barplot for injuries
barplot(injuries$INJURIES, names = injuries$EVTYPE, xlab = "", ylab = "Total Injuries", col = "orange", ylim= c(0, 26000), las=2, main = "Top 10 Weather Events Causing Injuries")

#Barplot for fatalities
barplot(fatalities$FATALITIES, names = fatalities$EVTYPE, xlab = "", ylab = "Total Deaths", col = "red", ylim= c(0, 3000), las=2, main = "Top 10 Weather Events Causing Death")
```

```{r fig.width=4, fig.heigh=10, echo=FALSE,out.extra='style="float:left"'}
op <- par(mar = c(12,6,4,2) + 0.1)
barplot(injuries$INJURIES, names = injuries$EVTYPE, xlab = "", ylab = "Total Injuries\n", col = "orange", ylim= c(0, 26000), las=2, main = "Events Causing Injuries")
par(op)
```

```{r, fig.width=4, fig.heigh=10, echo=FALSE, out.extra='style="float:left"'}
op <- par(mar = c(12,6,4,2) + 0.1)
barplot(fatalities$FATALITIES, names = fatalities$EVTYPE, xlab = "", ylab = "Total Deaths", col = "red", ylim= c(0, 3000), las=2, main = "Events Causing Death")
par(op)
```
<div style="clear: both;"></div>
Basing on the results we can say that Heat, Tornado and Flood cause the most deaths.Tornado, Lighting and Flood cause the most injuries.

Economical impact was sumarized in previous section, and the results are shown below. They answer on the second question.

```{r, echo=TRUE, cache=TRUE}
cropDmg
```

```{r, echo=TRUE, cache=TRUE}
propDmg
```

```{r, echo=TRUE, cache=TRUE}
totalDmg
```

```{r, echo=TRUE, fig.keep="none"}
#Barplot for crop damage
barplot(cropDmg$CROP_DMG/10^9, las = 3, names.arg = cropDmg$EVENT, main = "Greatest Crop Damage", ylab = "Cost of damages ($ billions)", col = "blue", ylim = c(0, 16))

#Barplot for property damage
barplot(propDmg$PROP_DMG/10^9, las = 3, names.arg = propDmg$EVENT, main = "Greatest Property Damage", ylab = "Cost of damages ($ billions)", col = "green", ylim = c(0, 200))

#Barplot for total damage
barplot(totalDmg$TOTAL_DMG/10^9, las = 3, names.arg = totalDmg$EVENT, main = "Greatest Total Damage", ylab = "Cost of damages ($ billions)", col = "violet", ylim = c(0, 200))
```

```{r fig.width=3, fig.heigh=10, echo=FALSE,out.extra='style="float:left"'}
op <- par(mar = c(12,4,4,2) + 0.1)
barplot(cropDmg$CROP_DMG/10^9, las = 3, names.arg = cropDmg$EVENT, main = "Greatest Crop Damage", ylab = "Cost of damages ($ billions)", col = "blue", ylim = c(0, 16))
par(op)
```

```{r, fig.width=3, fig.heigh=10, echo=FALSE, out.extra='style="float:left"'}
op <- par(mar = c(12,4,4,2) + 0.1)
barplot(propDmg$PROP_DMG/10^9, las = 3, names.arg = propDmg$EVENT, main = "Greatest Property Damage", ylab = "Cost of damages ($ billions)", col = "green", ylim = c(0, 200))
par(op)
```

```{r, fig.width=3, fig.heigh=10, echo=FALSE, out.extra='style="float:left"'}
op <- par(mar = c(12,4,4,2) + 0.1)
barplot(totalDmg$TOTAL_DMG/10^9, las = 3, names.arg = totalDmg$EVENT, main = "Greatest Total Damage", ylab = "Cost of damages ($ billions)", col = "violet", ylim = c(0, 200))
par(op)
```
<div style="clear: both;"></div>

Based on the plots above, we find that heat and flood couse the most crop damage, flood and wind the most property damage. Additionally, this two observations shows us that the greatest total damge couse flood & high surf and wind & storm.

## Conclusion
From all above data we can conclude that the highest impact on Us People health has Heat and Tornado, while Flood and High Surf causes the greatest total damage.